Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-22 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
more edits

Let us now consider the quartic EFT of GR, as our final example.
%
In Fig.~\ref{fig:qEFT_exec_sum} we show the posteriors on $\ell_{\rm qEFT}$ for GW150914 (top panel),
GW200129 (middle panel) for $N_{\rm max} = 0$, which are qualitatively similar to the cubic EFT of GR.
%
We find that both the EFT~\eqref{eq:eft_bound} and ParSpec~\eqref{eq:parspec_bound} are satisfied,
hence allowing us to infer bounds on $\ell_{\rm qEFT}$.
%
As Fig.~\ref{fig:qEFT_cdf} shows the CDFs for both events reach 90\% for $\gm = M_{\rm f}$ before $\varepsilon = 1/2$.
%
Consequently, from the joint posterior of $\ell_{\rm qEFT}$ we quote our final result,
%
\begin{equation}
    \ell_{\rm qEFT} \leqslant 51.3~\textrm{km} \quad \textrm{at 90\% credible level}\,.
\end{equation}
%
% \abcomm{Please update the paragraph below.}
In this theory, we have considered only $N_{\rm max} = 0$. We find that the
inclusion of the spin corrections yields waveforms that for $\ell_{\rm qEFT} \gtrsim 80$~km,
for which the ringdown portions of the larger than the inspiral-plunge part. \abcomm{The last sentence
has some problems in English. I do not udnerstand it.}
%
To avoid this issue we have lowered the value of $\ell^{\rm max}_{\rm qEFT}$, but by doing so we have obtained posteriors
which were essentially uniform, as our prior, and thus uninformative. Hence, we do not quote any results for $N_{\rm max} = 1$.
%
Table~\ref{tab:summary_qeft} summarizes our findings for the quartic EFT of GR.
%
\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/qeft_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function for the coupling constant $\ell_{\rm qEFT}$ for
GW150914 (top panel), GW200129 (middle panel) and combined events (bottom panel).
}
\label{fig:qEFT_exec_sum}
\end{figure}

@@

% -------------------------------------------------------------
% \hs{Try to draw some conclusions that apply to all theories.}
% -------------------------------------------------------------

We presented an unified framework that combines the ParSpec formalism introduced
to model deviations to the GR QNMs~\cite{Maselli:2019mjd} with the \pSEOB~waveform model~\cite{Brito:2018rfr,Ghosh:2021mrv}.
%
We showed with concrete examples, how theory-specific QNM calculations of
slowly-rotating BHs in modified gravity theories can be mapped onto the
non-GR parameters of the ParSpec formalism.
%
Put together this allowed us to use the \pSEOB~waveform model to test four
modified gravity theories (EdGB, dCS, cubic, and quartic EFTs of GR) using
observational data from the LVK events GW150914 and GW200129.
%
We found, in particular that, within the interpretation of our non-GR theories as EFTs and the
region of validity of the ParSpec framework, the coupling constant of dCS gravity is bound to
$\ell_{\rm dCS} \leqslant 34.5$~km, at 90\% credible level, when stacking the posteriors
of GW150914 and GW200129.
%
In contrast, we could not place any constraint of the coupling constants in EdGB gravity $\ell_{\rm EdGB}$.
%
This dichotomy between these theories has an analogy with works that considered
the inspiral part of the GW signal alone.
%
In that case, it is dCS gravity which cannot be constrained with current GW
events, while EdGB can, as shown in Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
\hscomm{I will add a sentence here on Refs.~\cite{Alexander:2017jmt,Loutrel:2018ydv,Loutrel:2018rxs}.}

Our work shows that there are theories of gravity (such as dCS) which can by-pass observational
constraints from the inspiral phase alone, {\it yet} would not, if we
analyze the ringdown.
%
We hope that our result thus serve as additional motivation for future work
modeling the late-inspiral, merger and ringdown of coalescing BHs in modified
gravity theories.

@@

\hscomm{TODO: say that we could try to expand the waveform model to include modification to
the inspiral part too; maybe through FTI~\cite{Mehta:2022pcn}?}

\hscomm{TODO: add that we could try to parametrize the final BH mass and spin estimates formulas we
use at the moment~\cite{Hofmann:2016yih}. Maybe comment that a possible first attempt could follow~\cite{Buonanno:2007sv},
as done in~\cite{Jai-akson:2017ldo,Li:2020wse}. Say that ultimately we will need to informed by NR in beyond-GR.}

\hscomm{I suspect that by construction the $N_{\rm max} = 1$ cases will often fail to satisfy our
perturbative bounds. If so, add explanation here.}

@@

>>>>

Let us now consider the quartic EFT of GR, as our final example.
%
In Fig.~\ref{fig:qEFT_exec_sum} we show the posteriors on $\ell_{\rm qEFT}$ for GW150914 (top panel),
GW200129 (middle panel) for $N_{\rm max} = 0$, which are qualitatively similar to the cubic EFT of GR.
%
We find that both the EFT~\eqref{eq:eft_bound} and ParSpec~\eqref{eq:parspec_bound} are satisfied,
hence allowing us to infer bounds on $\ell_{\rm qEFT}$.
%
As Fig.~\ref{fig:qEFT_cdf} shows the CDFs for both events reach 90\% for $\gm = M_{\rm f}$ before $\varepsilon = 1/2$.
%
Consequently, from the joint posterior of $\ell_{\rm qEFT}$ we quote our final result,
%
\begin{equation}
    \ell_{\rm qEFT} \leqslant 51.3~\textrm{km} \quad \textrm{at 90\% credible level}\,.
\end{equation}
%
% \abcomm{Please update the paragraph below.}
In this theory, we have considered only $N_{\rm max} = 0$.
%
\hs{We found that the inclusion of the spin corrections produced waveforms which for $\ell_{\rm qEFT} \gtrsim 80$~km
had a ringdown segment larger (sometimes seconds long) that the inspiral-plugne segment.}
% for which the ringdown portions of the larger than the inspiral-plunge part. \abcomm{The last sentence
% has some problems in English. I do not udnerstand it.}
%
To avoid this issue we have lowered the value of $\ell^{\rm max}_{\rm qEFT}$, but by doing so we have obtained posteriors
which were essentially uniform, as our prior, and thus uninformative. Hence, we do not quote any results for $N_{\rm max} = 1$.
%
Table~\ref{tab:summary_qeft} summarizes our findings for the quartic EFT of GR.
%
\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/qeft_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function for the coupling constant $\ell_{\rm qEFT}$ for
GW150914 (top panel), GW200129 (middle panel) and combined events (bottom panel).
}
\label{fig:qEFT_exec_sum}
\end{figure}

@@

We presented an unified framework that combines the ParSpec framework
to model deviations to the GR QNMs~\cite{Maselli:2019mjd} with the \pSEOB~waveform model~\cite{Brito:2018rfr,Ghosh:2021mrv}.
%
We showed with concrete examples, how theory-specific QNM calculations of
slowly-rotating BHs in modified gravity theories can be mapped onto the
non-GR parameters of the ParSpec formalism.
%
Put together this allowed us to use the \pSEOB~waveform model to test four
modified gravity theories (EdGB, dCS, cubic, and quartic EFTs of GR) using
observational data from the LVK events GW150914 and GW200129.

We found, in particular, that within the interpretation these theories as EFTs and the
region of validity of the ParSpec framework, the coupling constant of dCS gravity is bound to
$\ell_{\rm dCS} \leqslant 34.5$~km, at 90\% credible level, when stacking the posteriors
of GW150914 and GW200129.
%
In contrast, we could not place any constraint of the coupling constants in EdGB gravity $\ell_{\rm EdGB}$.
%
This dichotomy between the two theories has an analogy with works that considered
the inspiral part of the GW signal alone.
%
In that case, it is dCS gravity which cannot be constrained with current GW
events, while EdGB can, as shown in Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
% HS: changed my mind -- I think it would divert from the original flow.
% \hscomm{I will add a sentence here on Refs.~\cite{Alexander:2017jmt,Loutrel:2018ydv,Loutrel:2018rxs}.}
%
We hope that our results serve as further motivation to address these
challenging problems.

@@

% \hscomm{TODO: say that we could try to expand the waveform model to include modification to
% the inspiral part too; maybe through FTI~\cite{Mehta:2022pcn}?}

% \hscomm{TODO: add that we could try to parametrize the final BH mass and spin estimates formulas we
% use at the moment~\cite{Hofmann:2016yih}. Maybe comment that a possible first attempt could follow~\cite{Buonanno:2007sv},
% as done in~\cite{Jai-akson:2017ldo,Li:2020wse}. Say that ultimately we will need to informed by NR in beyond-GR.}

% HS: I did not add a comment on this, because of our discussion of how the N = 1 results will
% not be "permanent" as the N = 0 one, until we have QNM results valid up to ~ 0.7
% \hscomm{I suspect that by construction the $N_{\rm max} = 1$ cases will often fail to satisfy our
% perturbative bounds. If so, add explanation here.}

@@

Our work shows that there are theories of gravity (such as dCS) can be unconstrained by studying the
inspiral portion of GW events, {\it yet} can be constrained, if we analyze the ringdown.
%
Our results show that to fulfill the potential to test GR with GW observations,
further work on modeling of the late-inspiral-(plunge)-merger and ringdown of
coalescing BHs  (with NR simulations) and, complementary, the calculations of
QNMs of rapidly-rotating BHs in modified gravity theories (with BH perturbation
theory) are necessary.
